Modelling BCG vaccination in the UK

### 2019-12-11 rmds/04-tb-epi-england.Rmd
dodd minor corrections up to 6

For all outcomes considered I adjusted for the same set of demographic variables that were both highly complete and also plausibly linked to missingness for all outcomes considered. These were: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Complete case analysis has been used, with the dataset limited to notifications from 2010 and on-wards as socio-economic status was not collected prior to this.

@@

```{r bcgvacc-miss}
model_miss_results[["bcgvacc"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). Socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that BCG status is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r bcgvaccyr-miss}
model_miss_results[["bvcvaccyr"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(format = "latex",
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that year of BCG vaccination is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm")
```

@@

```{r dateofdeath-miss}
model_miss_results[["dateofdeath"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that there is some evidence that date of death is missing at random for ethnic group, with weaker evidence for all other variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r tomdeathrealat-miss}
model_miss_results[["tomdeathrealat"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that cause of death is missing at random for ethnic group and UK birth status, with little evidence for any other variables", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r symptonset-miss}
model_miss_results[["symptonset"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of symptom onset is missing not at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "HOLD_position")
```

@@

```{r datediag-miss}
model_miss_results[["datediag"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of diagnosis is missing at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r starttreatdate-miss}
model_miss_results[["starttreatdate"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. There is little evidence that the missing data for the date of starting treatment is associated with any variable considered, except for year of notification.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>%   
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r txenddate-miss}
model_miss_results[["txenddate"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of ending treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

An alternative measure is to use the date of symptom onset, this represents the closest approximation to the date when a case became infectious. Unfortunately there are multiple issues with this measure, the first of which being is that it is only `r symptonset_complete` complete across the data extract. Additionally, completeness changes with time, with  `r complete_sym_year$Complete[1]` complete in 2000,  `r complete_sym_year$Complete[9]` complete in 2008, and `r complete_sym_year$Complete[15]` complete in 2014. This could lead to spurious trends in the number of cases. Perhaps most importantly the date of symptom onset is highly susceptible to recall bias with the majority of cases becoming symptomatic on the first of each month (Figure \@ref(fig:plot-detection-sympton) d.)), with some evidence that a greater number of cases occur in January than would be expected (Figure \@ref(fig:plot-detection-sympton) c.)). Theses biases may also be the result of the defaults used during data entry with the first of the month or the first month of the year being used when the exact date is not known. Another possible measure of the number of cases is the date of diagnosis, this should be a more reliable variable than the date of symptom onset, as it does not rely on the recall of the case. However it is only `r datediag_complete` complete across the dataset, with strong evidence of increasing completeness going from  `r complete_diag_year$Complete[1]` complete in 2000, to `r complete_diag_year$Complete[15]` complete in 2014. This trend would be hard to properly account for in any analysis and therefore this variable should not be used as a primary measure.

@@

Both the mid-year ONS population estimates,[@storey2016] and the LFS are assessed for performance and quality elsewhere.[@NHSTraffordCCGGoverningBody2015; @Date2012] However, both have several failing that it is important to note, as they could introduce bias in future analysis. Whilst the ONS mid-year, and LFS estimates compare well when aggregated by age (Figure \@ref(fig:plot-overall-pop)) there is more disagreement when they are broken down by 5 year age groups (Figure \@ref(fig:plot-age-strat)). For those at working age both data sources are comparable (with approximately a 1% difference across all years). However, for children, young adults, and those who are 85+ the LFS underestimates the total population. This is particularly the case for older adults with between a 5% and 20% discrepancy for those aged 85-89 and a 25% to 45% discrepancy between those aged 90+. This could be problematic as these age groups often have the most severe outcomes to TB infection. A pragmatic approach to this is to exclude those aged 90+ from future analysis as results for this age group will be subject to large amounts of uncertainty which will be difficult to directly incorporate into the results.

@@

```{r age-dist-pop, fig.scap = "The estimate proprotion of the population in each 5 year age group stratified by UK birth status for 2000, 2008, and 2016.", fig.cap = "The estimate proprotion of the population in each 5 year age group stratified by UK birth status for 2000, 2008, and 2016.", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/age-dist-pop.png")
```

@@

Incidence has varied with time, increasing from `r IncTotalAllCases[[1]]` in 2000 to a maximum of `r IncTotalAllCases[[3]][2]` in `r IncTotalAllCases[[3]][1]`, since when it has decreased to a low of `r IncTotalAllCases[[2]]` in 2015 (Figure \@ref(fig:plot-overall-inc-rates)). This may indicate that TB control efforts are proving effective in preventing TB outbreaks, or may be driven by changes in the composition of those immigrating to England. It also highlights the lack of progress in reducing TB burden in England over the previous two decades, with little evidence of a decrease in overall incidence rates from 2000 until 2015. In the non-UK born incidence rates increased dramatically from 2000 to 2005, since when they have fallen consistently. In comparison, incidence rates fell in the UK born from 2000 until 2005 and then increased until 2012, since when they too have decreased year-on-year. This may indicate that  incidence rates in the two populations are linked, with incidence rates in the non-UK born driving incidence rates in the UK born with some time lag. Alternatively, it may be that incidence rates in the two populations are only weakly linked, or not linked at all. In this scenario the TB endemic in England would actually be two nearly separate endemics, each with different drivers. These scenarios can be differentiated using trends in age-specific incidence rates, and with statistical (Chapter \@ref(direct-eff))) and dynamic modelling (Chapter \@ref(model-development)).

@@

```{r plot-all-mort-age-dist, message=FALSE, warning=FALSE, fig.scap = "Age distribution (in 5 year age groups) of the case all-cause mortality rate presented on a square root scale.", fig.cap="Age distribution (in 5 year age groups) of the case all-cause mortality rate presented on a square root scale. Estimates are stratified by BCG and UK birth status. Point estimates and 95\\%  confidence intervals are shown for each case rate estimate. In both populaitons the case all-cause fatality rate increases with age, and has a secondary peak in early childhood (0-4). The all-cause case fatality rate is higher in BCG unvaccinated cases, compared to vaccinated cases, from early adulthood until 50 years of age in the UK born. There is less evidence of a difference in case fatality rates in the non-UK born. Case missing BCG status are more likely to die in both populations, with young non-UK born children being particularly at risk.", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-all-mort-age-dist.png")
```

@@

Similarly to all-cause deaths, deaths due to TB declined in the UK born over time but remained stable in the non-UK born (Figure \@ref(fig:plot-tb-mort-case-rate)). The case TB fatality rate also increased over time in both populations, with the rate again being higher in the UK born than in the non-UK born. There was still evidence of a higher case rate in those unvaccinated but the evidence for this was weaker. Comparing case TB fatality rates was difficult due to the large amount of uncertainty (Figure \@ref(fig:plot-tb-mort-age-dist)). However, there is some evidence to suggest that those missing BCG status, who were UK born, and those who were older were more likely to die from TB.

>>>>

For all outcomes considered I adjusted for the same set of demographic variables that were both highly complete and also plausibly linked to missingness for all outcomes considered. These were: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Complete case analysis has been used, with the dataset limited to notifications from 2010 and on-wards as socio-economic status was not collected prior to this.

@@

```{r bcgvacc-miss}
model_miss_results[["bcgvacc"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-45 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). Socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that BCG status is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r bcgvaccyr-miss}
model_miss_results[["bvcvaccyr"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(format = "latex",
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that year of BCG vaccination is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm")
```

@@

```{r dateofdeath-miss}
model_miss_results[["dateofdeath"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that there is some evidence that date of death is missing at random for ethnic group, with weaker evidence for all other variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r tomdeathrealat-miss}
model_miss_results[["tomdeathrealat"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that cause of death is missing at random for ethnic group and UK birth status, with little evidence for any other variables", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r symptonset-miss}
model_miss_results[["symptonset"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of symptom onset is missing not at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "HOLD_position")
```

@@

```{r datediag-miss}
model_miss_results[["datediag"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of diagnosis is missing at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r starttreatdate-miss}
model_miss_results[["starttreatdate"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. There is little evidence that the missing data for the date of starting treatment is associated with any variable considered, except for year of notification.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>%   
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r txenddate-miss}
model_miss_results[["txenddate"]] %>% 
  dplyr::select(-`P value (Wald)`) %>% 
  rlang::set_names(., nm = names(.) %>% stringr::str_replace_all("\\(LRT\\)", "")) %>% 
kable(
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of ending treatment as an outcome, adjusted for: year, sex, age (grouped as grouped as 0-14 year olds, 15-44 year olds, 45-64 year olds, and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

An alternative measure is to use the date of symptom onset. Unfortunately there are multiple issues with this measure, the first of which being is that it is only `r symptonset_complete` complete across the data extract. Additionally, completeness changes with time, with  `r complete_sym_year$Complete[1]` complete in 2000,  `r complete_sym_year$Complete[9]` complete in 2008, and `r complete_sym_year$Complete[15]` complete in 2014. This could lead to spurious trends in the number of cases. Perhaps most importantly the date of symptom onset is highly susceptible to recall bias with the majority of cases becoming symptomatic on the first of each month (Figure \@ref(fig:plot-detection-sympton) d.)), with some evidence that a greater number of cases occur in January than would be expected (Figure \@ref(fig:plot-detection-sympton) c.)). Theses biases may also be the result of the defaults used during data entry with the first of the month or the first month of the year being used when the exact date is not known. Another possible measure of the number of cases is the date of diagnosis, this should be a more reliable variable than the date of symptom onset, as it does not rely on the recall of the case. However it is only `r datediag_complete` complete across the dataset, with strong evidence of increasing completeness going from  `r complete_diag_year$Complete[1]` complete in 2000, to `r complete_diag_year$Complete[15]` complete in 2014. This trend would be hard to properly account for in any analysis and therefore this variable should not be used as a primary measure.

@@

Both the mid-year ONS population estimates,[@storey2016] and the LFS are assessed for performance and quality elsewhere.[@NHSTraffordCCGGoverningBody2015; @Date2012] However, both have several failings that it is important to note, as they could introduce bias in future analysis. Whilst the ONS mid-year, and LFS estimates compare well when aggregated by age (Figure \@ref(fig:plot-overall-pop)) there is more disagreement when they are broken down by 5 year age groups (Figure \@ref(fig:plot-age-strat)). For those at working age both data sources are comparable (with approximately a 1% difference across all years). However, for children, young adults, and those who are 85+ the LFS underestimates the total population. This is particularly the case for older adults with between a 5% and 20% discrepancy for those aged 85-89 and a 25% to 45% discrepancy between those aged 90+. This could be problematic as these age groups often have the most severe outcomes to TB infection. A pragmatic approach to this is to exclude those aged 90+ from future analysis as results for this age group will be subject to large amounts of uncertainty which will be difficult to directly incorporate into the results.

@@

```{r age-dist-pop, fig.scap = "The estimated proprotion of the population in each 5 year age group stratified by UK birth status for 2000, 2008, and 2016.", fig.cap = "The estimate proprotion of the population in each 5 year age group stratified by UK birth status for 2000, 2008, and 2016.", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/age-dist-pop.png")
```

@@

Incidence has varied with time, increasing from `r IncTotalAllCases[[1]]` in 2000 to a maximum of `r IncTotalAllCases[[3]][2]` in `r IncTotalAllCases[[3]][1]`, since when it has decreased to a low of `r IncTotalAllCases[[2]]` in 2015 (Figure \@ref(fig:plot-overall-inc-rates)). This may indicate that TB control efforts are proving effective in preventing TB outbreaks, or may be driven by changes in the composition of those immigrating to England. It also highlights the lack of progress in reducing TB burden in England over the previous two decades, with little evidence of a decrease in overall incidence rates from 2000 until 2015. In the non-UK born incidence rates increased dramatically from 2000 to 2005, since when they have fallen consistently. This may be driven by a change in the composition of the non-UK born population or it may be the result of increased screening of those entering the UK. In comparison, incidence rates fell in the UK born from 2000 until 2005 and then increased until 2012, since when they too have decreased year-on-year. This may indicate that  incidence rates in the two populations are linked, with incidence rates in the non-UK born driving incidence rates in the UK born with some time lag. Alternatively, it may be that incidence rates in the two populations are only weakly linked, or not linked at all. In this scenario the TB endemic in England would actually be two nearly separate endemics, each with different drivers. These scenarios can be differentiated using trends in age-specific incidence rates, and with statistical (Chapter \@ref(direct-eff))) and dynamic modelling (Chapter \@ref(model-development)).

@@

```{r plot-all-mort-age-dist, message=FALSE, warning=FALSE, fig.scap = "Age distribution (in 5 year age groups) of the case all-cause mortality rate presented on a square root scale.", fig.cap="Age distribution (in 5 year age groups) of the case all-cause mortality rate presented on a square root scale. Estimates are stratified by BCG and UK birth status. Point estimates and 95\\%  confidence intervals are shown. In both populaitons the case all-cause fatality rate increases with age, and has a secondary peak in early childhood (0-4). The all-cause case fatality rate is higher in BCG unvaccinated cases, compared to vaccinated cases, from early adulthood until 50 years of age in the UK born. There is less evidence of a difference in case fatality rates in the non-UK born. Case missing BCG status are more likely to die in both populations, with young non-UK born children being particularly at risk.", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-all-mort-age-dist.png")
```

@@

Similarly to all-cause deaths, deaths due to TB declined in the UK born over time but remained stable in the non-UK born (Figure \@ref(fig:plot-tb-mort-case-rate)). The case TB fatality rate also increased over time in both populations, with the rate again being higher in the UK born than in the non-UK born. There was still evidence of a higher TB related mortality rate in those unvaccinated but the evidence for this was weaker. Comparing case TB fatality rates was difficult due to the large amount of uncertainty (Figure \@ref(fig:plot-tb-mort-age-dist)). However, there is some evidence to suggest that those missing BCG status, who were UK born, and those who were older were more likely to die from TB.

```{r plot-tb-mort-case-rate, message=FALSE, warning=FALSE, fig.scap = "a.) Cases that died from TB by year of notification stratified by UK birth and BCG status, b.) Case TB fatality rate stratified by UK birth and BCG status.", fig.cap = "a.) Cases that died from TB by year of notification stratified by UK birth and BCG status, b.) Case TB fatality rate stratified by UK birth and BCG status. Point estimates along with 95\\% confidence are shown for all estimates. TB mortality has reduced over time in the UK born but remained stable in the non-UK born. This is also reflected in the case fatality rate with the UK born having a higher rate regardless of BCG status. The recording of BCG status has improved over time but it appears that for years with data BCG unvaccinated cases have a higher TB case fatality rate in both the UK and non-UK born. In both populations those missing UK birth status are more likely to die from TB. Data is incomplete for 2015, with cases that survived being potentially more likely to be missing than those that died. This may be the cause of the observed increase in uncertainty and may also have resulted in a biased TB fatality rate for 2015.", out.extra = ""}
knitr::include_graphics("chapters/tb-epi-england/figures/plot-tb-mort-case-rate.png")
```
